---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# yarrariver

<!-- badges: start -->
<!-- badges: end -->

yarrariver is a R package for exploring Yarra River water quality at CHANDLER HIGHWAY KEW, comparing the 1990s with recent years (≥ 2015).
It contains two cleaned datasets and a Shiny app that help you reproduce the core findings.

- `yarra_wq_period`: each row represents an observation of measurement labelled by `period`.
- `wq_hourly_median`: derived hourly summaries (median + IQR) from `yarra_wq_period`.
- `run_yarra_app()`: a helper to launch the Shiny app.

## Installation

You can install the development version of yarrariver from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("ETC5523-2025/assignment-4-packages-and-shiny-apps-KaiyiYe")
```

## Load the data

```{r eval=FALSE}
library(yarrariver)

data(yarra_wq_period)
data(wq_hourly_median)
```

## Example

To compare water quality over time, examine changes in recorded measurements. For example, check whether recent pH levels are higher or lower than in the 1990s. You can visualise this by plotting pH distributions by period：

```{r eval=FALSE}
ph_dist <- dplyr::filter(yarra_wq_period, parameter == "pH")

ggplot2::ggplot(
  ph_dist,
  ggplot2::aes(x = period, y = value, colour = period)
) +
  ggplot2::geom_boxplot(outlier.shape = NA, width = 0.65) +
  ggplot2::geom_jitter(width = 0.15, alpha = 0.45, size = 0.9) +
  ggplot2::labs(
    title = "Distribution by Period — pH",
    x = NULL, y = "Value"
  ) +
  ggplot2::theme_bw(base_size = 16) +
  ggplot2::theme(aspect.ratio = 1, legend.position = "none")
```

You can also explore changes interactively with the Shiny app—select a parameter or period to view its distribution or hourly medians:

```{r eval=FALSE}
run_yarra_app()
```

## Overview of data

- Source: [Victorian Water Measurement Information System (WMIS)](https://data.water.vic.gov.au/).

- Site coverage: CHANDLER HIGHWAY KEW.

- Parameter: pH, Water Temperature, Turbidity, Salinity as electrical conductivity.

You can find the data cleaning script in `data-raw/prepare_yarra_data.R`.

## Learn more

A vignette can help you learn more about this package.





